\documentclass[12pt,a4paper]{article}

\usepackage[T1]{fontenc}

\setcounter{secnumdepth}{-1}

\title{Pointer and Trail Sharing as Methods for Social Navigation}

\author{Eivind Uggedal\\
  \texttt{eivindu@ifi.uio.no}}

\date{}


\begin{document}

\maketitle{}

% TODO: REMEMBER TO USE CONCRETE EXAMPLES

The aim of this essay is to compare and contrast two papers in the field of
social navigation in the context of the web. First there will be summaries
of the papers, highlighting what I think are the most important aspects.
Then I'll compare the complementary aspects and contrast the contradictory
features of these two articles. Lastly I'll discuss my personal
opinions of the subject and how it all relates to the state of the modern web
infamously known as ``Web 2.0'' and finally come to a conclusion.

\section{Summary of Articles}

I decided to review two of the most cited papers in this field. While both of
them are nearly a decade old they both have important insights which could be
just as valuable today as when they were written.

\subsection{Supporting Social Navigation on the World Wide Web}
The first article \cite{dieberger97} describes navigational behavior on the
web, the state of social navigation support in the tools of that time and
reviews two prototype systems with social navigation features.

Dieberger is one of the first to compare social navigation in our concrete
real world to activities taking place on the Internet and particularly on the
World Wide Web. As he points out the term social navigation was coined as the
act when:
``\ldots movement from one item to another is provoked as an artifact of the
activity of another or a group of others.'' \cite{dourish94} Dieberger builds
on this definition and looks at the problem from a more practical standpoint.

\subsubsection{Navigational Behavior on the Web}

The author describes examples of social navigation observable on the web where
pointers characterized as URLs are exchanged and highlights the superiority of
such forms of navigation compared to aimlessly browsing the web. This sharing
of pointers to resources can be done deliberately by sending such
items trough interaction systems as e-mail both where the the receiver
actively requested them or just passing them around hoping that they
will be useful for the recipient. One can also take a more passive approach
where pointer pages are published on the web for everyone to benefit from in
their search for relevant information.

Users need a clear notion of structure for the available information due to
the Internets size and increasing growth. Diberger sees pointer collections
shared by both individuals and corporations as a solution to this
problem combined with readily available meta information of the webs
different kinds of resources. This purely social process as the author
describes it is important since it combats the tedious process of searching for
information and are uncolored of commercial interests.

\subsubsection{Support for Interaction on the Internet}

The majority of e-mail and newsgroup systems of the time had little support
for distinguishing URLs from other textual content making it hard for users to
easily share and use these elements. Chat systems offered sharing of URLs in
real time but also these systems offered unnatural handling of them.
Apart from verbosely talking about the resources there was not any real
interaction present. One could not for instance point out certain parts of a
document explains Dieberger.

There is also problems with usage of URLs in our daily world. Their rather
cryptic nature and their length makes it hard to remember them without noting
them down. Advances are being made in browsers so that the protocol (HTTP)
and sometimes the sub-domain (typically \texttt{www}) and top level domain can
be omitted from the URL when typing them in.

Dieberger makes the case for hiding URLs from the end user based on the
inherent problems with their format and how unnatural tools of the that time
supported them. Users have a need to point out entities of information in an
information system and this interaction can be made simpler and more efficient
if they can ignore the address of the resource but share and access it trough
a handle, a metaphor wrapped around the complex URL as the author calls it.
This handle would also need to provide easily accessible meta information
about the resource it represents so that users can make an informed decision
about whether the document is worth visiting without actually visiting and
reviewing the document itself. Lastly this page handle have to be an
integrated part of the information system it's implemented in so that it
can function as unobtrusive as possible.

Awareness of other users on the web can create opportunities for several forms
of social interaction forms and thereby forms of social navigation. Dieberger
says that this perception of other individuals acts as a special type of meta
information giving feedback on where people are going, in what amounts and
who they are. It's important to be aware of the privacy issues that arises
when leveraging such information directly. By using hints about other users'
behavior indirectly one can follow a safer path which eliminate such
concerns. Nevertheless the communication systems of this time presents the web
as a vast information repository where you're the only user, not visualizing a
shared perspective where individuals can take advantage of each other's
actions.

\subsubsection{Examples of Social Navigation Systems}

The author uses two prototypes of information systems to highlight what he
sees at the defining aspects of social navigation on the web. First we have
the ``Juggler'' system, a text based real-time communication system
with an integrated web proxy doubling as an alternative communication
interface and historical repository. It's most evident features with
regard to social navigation support include several techniques for handling
exchange of web pointers as seamless as possible and methods of visualizing
historical aspects of information.

With the extensions made to this system it was possible for users to share a
web page they were currently viewing with the push of a single button. The
receiver(s) would then be presented with a new browser window of that exact
page automatically. The author concludes that this simple way for users to
point out web pages increased the social navigation activity.
This supports his postulation of users unwillingness to have URLs made visible
for them.

The system incorporate the idea of ``hit-counters'' on web pages
to make visible the frequency of usage for it's various navigation paths. This
turns the system into a \emph{history-enriched} \cite{hill94}
environment where so called
\emph{read wear} \cite{hill92} can be seen which points users in the direction
of the most visited and used information entities. The textual communication
system uses verbal descriptions to implement read wear, denoting both how often
read-only sources are visited and also the frequency of postings for editable
information. The author emphasizes the importance of decaying these
representations making recent information more important than older
information. The notion of read wear were taken even further in the web
gateway visualizing frequency of usage trough color coding objects.

An alternative system to the conventional browser bookmark list
named ``Vortex'' is also used to highlight topics of social navigation. The
prototypical implementation handles URLs and associated meta-data (usage
statistics, keywords, summary etc.) as a coherent object. These objects,
referred to as handles by the author, can also be annotated so that it's
perceivable for the author which handles are the most important trough the
size of it's associated icon. Even though this system was not intentionally
made for communication and collaboration purposes such features have been
added in the form of automatic functions for copying whole pointer objects
into emails with their meta-information and parsing of URLs in incoming email
into handles. Dieberger also envisions a port of the system to a more system
independent applet platform enabling users to export their handle collections
to a standard markup format, upload it to a central server, enabling other
users to contribute to this shared collection.

\subsection{Footprints: History-Rich Tools for Information Foraging}

The second article \cite{wexelblat99} communicates the authors proposed theory
of interaction history, it's application in real world tools, and the results
of controlled usage studies. 

Wexelblat et al. contrasts the digital world of computers against our
non-digital everyday world with respect to the formers lack of history. In our
traditional world we exploit such historical information traces:
``\ldots to guide our actions, to make choices, and to find things of
importance or interest.'' \cite{wexelblat99}
Objects that contain information that make such interactions possible are
described as \emph{history-rich} according to the authors. It's important to
note that usage of interaction history in the physical world are mostly
achieved without any concious efford and therefore feels natural to us.

This apparent lack of history in our digital systems must be sorted out such
that future users can take advantage of past users historical traces left when
they were working on solving problems similar to the current user's. This
should make problem solving easier and more efficient by recovering past
findings and avoiding mistakes made in the past. Without such systems users
have to behave like they are the first and only people using the information
-- everyone have to start from scratch in their problem solving tasks.

\subsubsection{Theoretical Interaction History Framework}

Wexelblat et al. have taken earlier insights into history-rich objects and
their usage \cite{hill92, hill94} and extended it into a theoretical framework
consisting of six properties, all characterizing interaction history based
systems:

\begin{description}
  \item[Proxemic versus distemic] describe how close people and places are
    related. A \emph{proxemic} space is transparent to it's users since it's
    signs and structures are easily comprehended. \emph{Distemic} places are on
    the other hand opaque to it's users. Translating this to interaction
    history systems means that a proxemic system relates well to it's users
    and takes advantage of past interaction experiences and knowledge.
  \item[Active versus passive] characterize the way interaction history is
    recorded and made available. One can leave records actively (saving a
    bookmark) or passively (a web client's history records). The latter
    involves no
    concious effort and is according to Wexelblat et al. therefore the desired
    kind as users problem solving processes remains uninterrupted.
  \item[Rate/form of change] explains how history builds up trough an
    \emph{accretion} process as more and more interactions take place. History
    also fades out as time goes by and more history accumulates. It's important
    to deal with these processes so that all available information can be
    summarized in a way that makes efficient usage and access to it possible.
  \item[Degree of permeation] detail to what extent the interaction history is
    part of the hisotry-rich object itself, ranging from inseparable to
    completly separable.
    %TODO: MABY ELABORATE ON THIS A BIT
  \item[Personal versus social] means that history either can be social of
    nature (a groups shared bookmark collection) or personal (bookmark
    collection exclusive for one individual). Wexelblat et al. feels that
    the social kind of interaction history is more valuable since most problem
    solving tasks are collaborative in nature. The authors is stressing that
    the main advantage of interaction history lies in the way it enables
    newcomers to benefit from work completed in the past.
    
  \item[Kind of information] refers to the idea that there exists infinite
    kinds of information history. The task that the users are trying to
    achieve determines what kinds of information that is of importance. The
    authors categorizes information into categories based on different uses of
    interaction history:
  \begin{description}
    \item[What] was done. Helpfull if users are trying to find usefull
      information, need reassurance or need guidance.
    \item[Who] has done something. Crucial for sociability, authority and
      authenticity. 
    \item[Why] something was done. Important for knowing whether the past users
      had similar intentions and for explanation and learning.
    \item[How] something was done. Essential for achieving naturalness and
      transparency.
  \end{description}
\end{description}

\subsubsection{Interaction History on the Web}

The ``Footprints'' system is described by the authors as an attempt to find
the valuable aspects of interaction history in our real world and how one can
capture history in digital systems for such applications. They wanted to
validate their theoretical framework and did so by applying this theory to the
problem of how to navigate a complex information space as the web. The system
is an active navigational aid both visualizing interaction history
information and also enabeles users to navigate this information. If the user
knows what he want to find Footprints can help them find their way and
explain what they've found.

The system consists of several tools integrated into a normal web browser,
using navigation transitions as their fundamental information for historical
preservence. All tools are based on a navigational metaphor:

\begin{description}
  \item[Maps] are used to show traffic trough the currently visited
    web site. Transitions from this page to other documents are visualized as
    links between several nodes. Read wear is existant trough visualization of
    a documents popularity as shades of different colors.
  \item[Trails] are ``\ldots coherent sequences of nodes followed by an
    individual'' \cite{wexelblat99}. Trails are visualized trough a path
    interface which acts as a low level view compared to the high level nature
    of the map interface. These trails are automatically recorded and the
    system useses several properties to distinguish between different paths.
    Trails are also visualized by their degree of usage, the ticker the links
    between nodes the higher the actual usage of the trail.
  \item[Signposts] are a way for users to comment on the interaction history
    they are using, giving feedback to both individual pages and trails.
\end{description}

Lastly users are given information about the usage of inline links of web
pages trough the use of textual annotations showing the percentage of users
who followed each particular link.

\subsubsection{Usage of Interaction History Tools}

The Footprints application was tested in a controlled environment by
Wexelblat et al. where several users were given the task of locating a used
car within a given price range. One group used normal web clients while
another group used the Footprints application.
Their pre-test hypoteses were that this tool
would increase the number of a alternative car dealerships found and reduce
the number of pages visited while browsing for these. The authors also had
ambitions of the tools ability to ease the task of locating
information and understanding discovered information in addition to
increasing the level of satisfaction among users.

The experiment only parially supported the authors claims. No notable
diffrences between number of alternatives between the two groups were found.
But the group using the history-enriched system reported signifficantly lower
values of their mean page count for completing their task.

With regards to subjective responses the authors found that experienced users
(users who knew the problem domain to some extent) could to a larger degree
take advantage of interaction history in their browsing. Novice users on the
other hand was actually less satisfied with using interation models than
normal browsing methods. Wexelblat et al. explains that experienced users
could make much better use of Footprint's representations since they had a
clear mental model of car information and thus making the application's
interaction history more proxemic.

During this experiment the authors found three different patterns of use for
Footprints. Some wandered down paths not used before within the system and
therefore received little help from the system. Others started out using
information within Footprints and took off in other directions when their
taste in cars started to differ from past users. Lastly it was observed that
some users started out without using information from Footprints, but when
their browsing brought them into trails of past users they seemed to take
advantage of their earlier work.

\section{Comparison of Articles}

Both articles describes how normal browsing behaviour on the web can be
improved by means of social navigation examplified in several tools. They take
very different approaches to include sociability in the web. Dieberger focues
on improving patterns of social navigation that have widespread use while
Wexelblat et al. have created tools which support navigating the web in
entirely new ways.

Dieberger describes tools where users actively shares information trough a
social process and then actively decides wheter to use this information.
Wexelblat et al. on the other hand emphasises on passively collecting
information that can be used actively by other users at a later stage. The
distinction between these two concepts can be examplified as the difference
between asking or beeing told what direction to take in a myriad of forest
trails and following the trail that is most worn out. In the first case the
sender have to exlicitly tell the receiver where to go, requiring concious
effort. This is contrasted in the last example where several users of the
forest leave traces of their action unconciously and without no additional
effort.

The behaviour of people's structuring of the web trough making structured
pointer pages available as described by Dieberger also applies to the
Footprints system. When people passively creates surfing trails and then
actively comment on both trails and individual documents they too are opposing
structures on the web.

In the Juggler system people are aware of eachother and can to a certain
degree know their identity as opposed to the Footprints system where all users
behaviour are merged and therefore making this an anonymous system. The
advantage of Diebergers approach is that you by using the system can learn
wich users to trust, using information provided by reputable users
before unknown people. Since the Footprints system tracks every movment of
it's users such an approach would bring forth a host of privacy issues.
Wexelbalt et al. states that this is a deliberate trade-off, they wanted to
focus on interaction history and not methods for personal privacy.

As discussed earlier Dieberger proved his thesis that users do not want to see
URLs while he was working on the Juggler system (extended into his work on the
Vortex system). This same idea is implemented
troughout the Footprints system of Wexelblat et al
trough not explicitly stated in the article, but evident in the design
of the system. Individual documents are denoted by their page title in all of
Footprint's tools and the URL is only part of the browsers location bar.

Usage of read- and edit-wear are apparent in both the Juggler and the
Footprints application. Both uses colorized visualizations and textual
representations where either is more appropriate. There seem to be diffrences
in what colors are used for displaying read- and edit wear. It could be
confusing for users when such presentation methods differ amongst different
applications and would therefore be advisable to either trying to establish
standard patterns or take advantage of conventions found in the wild, both on
the web and in our real world. Developing broad conventions is difficult as
diffrences among cultures also have to been taken into concideration.

\section{Contrasting}

Dieberger cites Dourish (citeme) as the originator of the term social
navigation.

\section{Discussion}

This section will try to highlight the state of social navigation support in
todays applications and the web in areas of pointer sharing and interaction
history.

\subsection{Dieberger}

\subsection{Sharing of Pointers}

\subsubsection{URL Recognition}

Dieberger describes the various problems with URL handeling in the systems of
that time and that this affects how likely people are to share pointers. Today
most email clients makes URLs clickable in contrast with similar systems a
decade ago. I surveyed some of the most popular web-based clients and also
clients for Microsoft Windows, Apple OS X and Linux. The results can be seen
in Table~\ref{table:email} (p.~\pageref{table:email}).

\begin{table}[h!b!p!]
  \caption{URL Recognition Support in Email Clients}
  \label{table:email}
  \begin{center}
    \begin{small}
      \begin{tabular}{l|ccc}
        \hline
        &
        \texttt{domain.tld} &
        \texttt{www.domain.tld} &
        \texttt{http://domain.tld} \\
        \hline
        \hline
        Google Gmail (2007-06) & $\surd$ & $\surd$ & $\surd$ \\
        Yahoo Mail (2007-06)   &         & $\surd$ & $\surd$ \\
        MS Hotmail (2007-06)   &         &         & $\surd$ \\
        SquirrelMail (1.4.10a) &         &         & $\surd$ \\
        Moz Thunderbird (1.0)  &         & $\surd$ & $\surd$ \\
        Evolution (2.0.2)      &         & $\surd$ & $\surd$ \\
        KMail (1.7.1)          &         & $\surd$ & $\surd$ \\
        Apple Mail (2.1.1)     &         &         & $\surd$ \\
        MS Outlook (2007)      &         & $\surd$ & $\surd$ \\
        \hline
      \end{tabular}
    \end{small}
  \end{center}
\end{table}

Note that this is when URLs are in plain text and not marked up as
HTML anchors. As one can see from the list all clients supported links with
protocol (HTTP) prefix. Two thirds of the clients supported links without
protocol prefix but with \texttt{www} prefix. The Google Gmail web-based
client stood out from the crowd with support from all kinds of valid URLs
without protocol prefix.

Compared to the days when Dieberger wrote his article the state of automatic
URL recognition is vastly improved even though some email clients could try to
mimic features of state of the art clients as Google Gmail. 

% POINTER PAGES => delicious


The pointer pages of today is del.icio.us in many ways. More socially
oriented. Aslo ``elswhere'' pages on several blogs functions as mass-pointer
pages.

Delicious pages have more links than personal pointer pages. Yahoo et al. was
organized hierarchical but such a system is hard to implement. Need to be
librarian. Some pointers match several categories. The solution: folsonomy,
tagging.

Bookmark collections have been improved in some browsers. Flock and FF with
extension integrates with del.icio.us. New firefox bookmarks in 3.0 are called
places, uses SQLite backend. Research this more.

Pointers and pointer collections are not handled as transparently as in the
Vortex system though. Improvments have been made in the OSes (OSX, Win, Gnome,
KDE DEs) to make links clickable and sendable by mail etc.

Supposed improvments to Vortex for building pointer collections
collaboratively is seen in collaborative tagging systems of today. ref
article.

Taking notes not supported at that time. See what various post-it like
extensions to the web, both server side and client side trough extensions, are
toing today and what it means.

No google at the time Dieberger says that finding information is almost
impossible. Research if search engines (spidered) was available or if only
directories as Yahoo were. ALTAVISTA/GOOGLE WAS PARTIALLY AVAILABLE AT THIS
TIME.

Lack of meta-information on the web. We have RDF, but no widespread use.
Microformats as a pragmatic grasroots kind of thing are taking off in some
communities.


\subsection{Wexelblat}

Wexelblat vs Trails.app: discuss the wanted addition of a purpose part of the
Footprints and se if it's implemented in Trails.app.

Trails.app is active for the users creating trails (check this, but think so)
but Footprints is passive (except for signpost/comment making).

\section{Conclusion}

Some smart conclusion.

\section{}


\bibliographystyle{apalike}
\bibliography{citations}

\end{document} 
